---
title: "Introduction to `ordbetareg`"
author: "Robert Kubinec"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{package_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


```{r setup, include = FALSE}

library(ordbetareg)
library(dplyr)
library(ggplot2)
library(haven)
library(brms)
library(tidyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```


# Overview

This notebook contains instructions for running the ordered beta regression model in the R package `ordbetareg`. `ordbetareg` is a front-end to `brms`, a very powerful regression modeling package based on the [Stan Hamiltonian Markov Chain Monte Carlo](https://mc-stan.org/) sampler. I only show in this vignette a small part of the features which are available via `brms`, and I refer the user to the [copious documentation with the package](https://paul-buerkner.github.io/brms/). Suffice it to say that most kinds of regression models can be fit with the software, including hierarchical, dynamic, nonlinear and multivariate models (are all of the above in combination). The `ordbetareg` allows for all of these features to be used with the ordered regression model distribution by adding this distribution to `brms`.

The ordered beta regression model is designed explicitly for data with upper and lower bounds, such as survey slider scales, dose/response relationships, and indexes. I refer you to a paper on the model if you are not familiar with it: https://osf.io/preprints/socarxiv/2sx6y/.

If you use the model in a paper, please cite it as:

Kubinec, Robert. "Ordered Beta Regression: A Parsimonious, Well-Fitting Model for Continuous Data with Lower and Upper Bounds." **Political Analysis.** 2022. Forthcoming.

```{r runmod}

# whether to run models from scratch
run_model <- F

```

# Define Custom Family

If you prefer to run `brms` directly rather than use this package, you can run the underlying code via the R script `define_ord_betareg.R` in the [paper Github repo](https://github.com/saudiwin/ordbetareg). Please not that I cannot offer support for this alternative, although it should work.

# Data Preparation

First, I load data from a Pew Forum survey that asked a question about respondents' views towards college professors.

```{r load_data}

data("pew")

pew %>% 
  ggplot(aes(x=as.numeric(therm))) +
  geom_histogram(bins=100) +
  theme_minimal() + 
  theme(panel.grid=element_blank()) +
  scale_x_continuous(breaks=c(0,25,50,75,100),
                     labels=c("0","Colder","50","Warmer","100")) +
  ylab("") +
  xlab("") +
  labs(caption=paste0("Figure shows the distribution of ",sum(!is.na(pew$therm))," non-missing survey responses."))

```

The distributions of feelings towards college professors contains both degenerate (0 and 100) and continuous responses between 0 and 100. To model it, the outcome needs to be re-scaled to lie strictly between 0 and 1. However, it is not necessary to do that ahead of time as the `ordbetareg` package will do that re-normalization internally. I also do some other data processing tasks:

```{r munge_data}

model_data <- select(pew,therm,age="F_AGECAT_FINAL",
                        sex="F_SEX_FINAL",
                        income="F_INCOME_FINAL",
                        ideology="F_IDEO_FINAL",
                        race="F_RACETHN_RECRUITMENT",
                        education="F_EDUCCAT2_FINAL",
                     region="F_CREGION_FINAL",
                        approval="POL1DT_W28",
                       born_again="F_BORN_FINAL",
                       relig="F_RELIG_FINAL",
                        news="NEWS_PLATFORMA_W28") %>% 
    mutate_at(c("race","ideology","income","approval","sex","education","born_again","relig"), function(c) {
      factor(c, exclude=levels(c)[length(levels(c))])
    }) %>% 
    # need to make these ordered factors for BRMS
    mutate(education=ordered(education),
           income=ordered(income))

```

The completed dataset has `r nrow(model_data)` observations. 

# Run In `ordbetareg`

Given these new functions, we can then run a `ordbetareg` model as usual. The `ordbetareg` function will take care of normalizing the outcome and adding additional information necessary for to estimate the distribution. However, any additional arguments can be passed to the underlying `brm` function. For example, in the code below I use the `backend="cmdstanr"` argument to `brm()`, which allows me to use the R package `cmdstanr` for estimating models. `cmdstanr` tends to have the most up to date version of Stan, though you must install it yourself. 

What you need to pass to the `ordbetareg` function are the standard components of any R model: a formula and a dataset that has all the variables mentioned in the formula. There are additional parameters that allow you to modify the priors, such as for the dispersion parameter `phi`. While in most cases these priors are sensible defaults, if you have data with an unusual scale (such as very large or very small values), you may want to change the priors to ensure they are not having an outsize effect on your estimates.

If you want to use some of `brms` more powerful techniques, such as multivariate modeling, you can also pass the result of a `bf` function call to the `formula` argument. I refer you to the `brmsformula()` [function help](https://paul-buerkner.github.io/brms/reference/brmsformula.html) for more details.

To demonstrate some of the power of using `brms` as a regression engine, I will model education and income as ordinal predictors by using the `mo()` function in the formula definition. By doing so, we can get a single effect for education and income instead of having to use dummies for separate education/income categories. As a result, I can include an interaction between the two variables to see if wealthier more educated people have better views towards college professors than poorer better educated people. Finally, I include varying (random) census region intercepts.

```{r run_brms}

if(run_model) {
  
  ord_fit_mean <- ordbetareg(formula=therm ~ mo(education)*mo(income) +
                               (1|region), 
                       data=model_data,
                cores=2,chains=2,
                refresh=0,
                # NOTE: remove these two options if you don't have 
                # cmdstanr set up (or you don't have 5 cores to burn per chain)
                threads=threading(5),
                backend="cmdstanr")
  
} else {
  
  data("ord_fit_mean")
  
}
```

The one divergent transition referenced above is due to the well-known funnel problem of the variance of the random intercepts, and I will ignore it for the purposes of this vignette.

# Post-Estimation

The first thing we can do is extract the model cutpoints and overlay them on the empirical distribution to see how the model is dividing the outcome into discrete-ish categories. We have to do transformation of the cutpoints using the inverse logit function in R (`plogis`) to get back values in the scale of the response, and I have to exponentiate and add the first cutpoint to get the correct value for the second cutpoint:

```{r plot_cut}

all_draws <- extract_draws(ord_fit_mean)

cutzero <- plogis(all_draws$dpars$cutzero)
cutone <- plogis(all_draws$dpars$cutzero + exp(all_draws$dpars$cutone))

pew %>% 
  ggplot(aes(x=therm)) +
  geom_histogram(bins=100) +
  theme_minimal() + 
  theme(panel.grid=element_blank()) +
  scale_x_continuous(breaks=c(0,25,50,75,100),
                     labels=c("0","Colder","50","Warmer","100")) +
  geom_vline(xintercept = mean(cutzero)*100,linetype=2) +
  geom_vline(xintercept = mean(cutone)*100,linetype=2) +
  ylab("") +
  xlab("") +
  labs(caption=paste0("Figure shows the distribution of ",sum(!is.na(pew$therm))," non-missing survey responses."))


```

We can see in the plot above that the model does a good job isolating values that are very close to the degenerate bounds of 0 and 100 from values that are more continuous in nature. 

We can plot the full predictive distribution relative to the original outcome. The model can't capture all of the modality in the distribution -- there are effectively four separate modes -- but it is reasonably accurate over the middle responses and the responses near the bounds.

```{r post_predict}

pp_check(ord_fit_mean) + theme_minimal()

```

We can see the coefficients from the model in table formusing the `modelsummary` package, which has reasonable (though not perfect) support for `brms` models. We'll specify to include only confidence intervals as other frequentist statistics have no Bayesian analogue (i.e. p-values). We'll also specify only the main effects of the ordinal predictors, and give the more informative names. 

```{r coef_plot}

modelsummary(ord_fit_mean,statistic = "conf.int",
                          metrics = "RMSE",
                          coef_map=c("bsp_moeducation"="Education",
                                     "bsp_moincome"="Income",
                                     "bsp_moeducation:moincome"="EducationXIncome"))
```

`modelsummary` tables have many more options can be output to both html and latex formats. 

# Marginal Effects

There is a related package, `marginaleffects`, that allows us to convert these coefficients into more meaningful marginal effect estimates, i.e., the effect of the predictors expresses as the actual change in the outcome on the 0 - 1 scale. We can't calculate marginal effects for all of the predictors as some of them are cutpoints, etc. (the function will return an error about the intercept), so we'll focus on one of the main effects we are interested in as in the table above:

```{r marg_effect}

library(marginaleffects)

marg_effs <- marginaleffects(ord_fit_mean,
                             variables="education")

summary(marg_effs) %>% knitr::kable()

```

We can see that we have separate marginal effects for each level of education due to modeling it as an ordinal predictor. As we can see with the raw coefficient in the table above, the marginal effects are all positive, though the magnitude varies across different levels of education.

# Understanding Clustering/Polarization of Respondents

As I explain in the paper, one of the main advantages of using a Beta regression model is its ability to model the dispersion among respondents not just in terms of variance (i.e. heteroskedasticity) but also the shape of dispersion, whether it is U or inverted-U shaped. Conceptually, a U shape would imply that respondents are bipolar, moving towards the extremes. An inverted-U shape would imply that respondents tend to cluster around a central value. We can predict these responses conditionally in the sample by adding predictors for `phi`, the scale/dispersion parameter in the Beta distribution. Higher values of `phi` imply a uni-modal distribution clustered around a central value, with increasing `phi` implying more clustering. Lower values of `phi` imply a bi-modal distribution with values at the extremes. Notably, these effects are calculated independently of the expected value, or mean, of the distribution, so values of `phi` will produce different shapes depending on the average value. 

The one change we need to make to fit this model is to add a formula predicting `phi` in the code below. Because we now have two formulas--one for the mean and one for dispersion--I use the `bf` function to indicate these two sub-models. I also need to specify `phi_reg` to be TRUE because some of the priors will change. 

Because there is no need to model the mean, I leave the first formula as `therm ~ 0`. I don't specify 1 for an intercept because the ordered regression model has other intercepts (i.e., the cutpoints). I then specify a separate model for `phi` with an interaction between `age` and `sex` to see if these covariates are associated with dispersion.


```{r run_brms_phi}

if(run_model) {
  
  ord_fit_phi <- ordbetareg(bf(therm ~ 0, 
                phi ~ age + sex),
                phi_reg = T,
                data=model_data,
                cores=2,chains=2,
                refresh=100,
                # again, remove these options if you don't have cmdstanr
                # installed or don't want to use multiple cores
                threads=threading(5),
                backend="cmdstanr")
  
} else {
  
  data("ord_fit_phi")
  
}


```

We can quickly examine the raw coefficients:

```{r phicoef}

summary(ord_fit_phi)

```

However, these are difficult to interpret as they relate to the Beta distribution, which is highly nonlinear. Generally speaking, higher values of `phi` mean the distribution is more concentrated around a single point. Lower values imply the distribution is more dispersed to the point that it actually becomes bi-modal, meaning that responses could be close to either 0 or 1 but are unlikely in the middle.

Because `phi` is a dispersion parameter, by definition the covariates have no effect on the average value. As a result, we'll need to use the `posterior_predict` function in `brms` if we want to get an idea what the covariates do. We don't have any fancy packages to do this for us, so we'll have to pass in two data frames, one with sex equal to female and one with sex equal to male. We'll want each data frame to have each unique value of age in the data. 

```{r plot_phi_sim}

# we can use some dplyr functions to make this really easy

female_data <- distinct(model_data, age) %>% 
  mutate(sex="Female")

male_data <- distinct(model_data, age) %>% 
  mutate(sex="Male")

to_predict <- bind_rows(female_data,
                        male_data) %>% 
  filter(!is.na(age))

pred_post <- posterior_predict(ord_fit_phi,
                               newdata=to_predict,
                               draw_ids=c(1:100))

tibble(phi_small=rbeta_mean(10000,mean(model_data$therm,na.rm=T),2),
       phi_big=rbeta_mean(10000,mean(model_data$therm,na.rm=T),3)) %>% 
  gather("phi_type","simulated_value") %>% 
  ggplot(aes(x=simulated_value)) +
  geom_density(alpha=0.5,aes(fill=phi_type)) +
  theme(panel.background = element_blank(),
        panel.grid=element_blank())

```

We can see that the `phi_big` distribution is more clustered around a central value while `phi_small` shows more movement towards the extremes of the distribution. However, the movement is modest, as the value of the coefficient suggests.

# Simulate Data

Finally, we can also simulate data from the ordered beta regression model with the `sim_ordbeta` function. This is useful either for examining how different parameters interact with each other, or more generally for power calculation by iterating over different possible sample sizes. I demonstrate the function here, though note that the vignette loads saved simulation results unless `run_model` is set to TRUE. Because each simulation draw has to estimate a model, it can take some time to do calculations. Using multiple cores a la the `cores` option is strongly encouraged to reduce processing time. 

To access the data simulated for each run, the `return_data=TRUE` option can be set. To get a single simulated dataset, simply use this option combined with a single iteration and set of parameter values. The data are saved as a list in the column `data` in the returned data frame. The chunk below examines the first 10 rows of a single simulated dataset (note that the rows are repeated `k` times for each iteration, while there is one unique simulated dataset per iteration). Each predictor is listed as a `Var` column from 1 to `k`, while the simulated outcome is in the `outcome` column.

```{r check_data}

  single_data <- sim_ordbeta(N=100,iter=1,
                             return_data=T)

# examine the first dataset

knitr::kable(head(single_data$data[[1]]))

```

By default, the function simulates continuous predictors. To simulate a binary variables, such as in a standard experimental design, use the `beta_type` function to specify `"binary"` predictors and `treat_assign` to determine the proportion assigned to treatment for each predictor. For a standard design with only one treatment variable, we'll also specify that `k=1` for a single covariate. Finally, to estimate a reasonable treatment effect, we will specify that `beta_coef=0.5`, which equals an increase of .5 on the logit scale. While it can be tricky to know a priori what the marginal effect will be (i.e., the actual change in the outcome), the function will calculate marginal effects and report them, so you can play with other options to see what gets you marginal effects/treatment effects of interest.

```{r sim_data_full}

if(run_model) {
  
  sim_data <- sim_ordbeta(N=c(250,500,750),
                          k=1,
                          beta_coef = .5,
                          iter=100,cores=10,
                          beta_type="binary",
                          treat_assign=0.3)
  
} else {
  
  data("sim_data")
  
}


```

For example, in the simulation above, the returned data frame stores the true marginal effect in the `marg_eff` column, and lists it as `r round(sim_data$marg_eff[1], 3)`, which is quite a large effect for a $[0,1]$ outcome. The following plot shows some of the summary statistics derived by aggregating over the iterations of the simulation. Some of the notable statistics that are included are power (for all covariates $k$), S errors (wrong sign of the estimated effect) and M errors (magnitude of bias). As can be seen, issues of bias decline markedly for this treatment effect size and a sample of 500 or greater has more than enough power.

```{r sim_plot}

sim_data %>% 
    select(`Proportion S Errors`="s_err",N,Power="power",
         `M Errors`="m_err",Variance="var_marg") %>% 
  gather(key = "type",value="estimate",-N) %>%
  ggplot(aes(y=estimate,x=N)) +
  #geom_point(aes(colour=model),alpha=0.1) +
  stat_summary(fun.data="mean_cl_boot") + 
  ylab("") +
  xlab("N") +
  scale_color_viridis_d() +
  facet_wrap(~type,scales="free_y",ncol = 2) +
  labs(caption=stringr::str_wrap("Summary statistics calculated as mean with bootstrapped confidence interval from simulation draws. M Errors  and S errors are magnitude of bias (+1 equals no bias) and incorrect sign of the estimated marginal effect respectively. Variance refers to estimated posterior variance (uncertainty) of the marginal effect(s).",width=50)) +
  guides(color=guide_legend(title=""),
         linetype=guide_legend(title="")) +
  theme_minimal() +
  theme(plot.caption = element_text(size=7),
        axis.text.x=element_text(size=8))

```

While the `sim_ordbeta` function has the ability to iterate over `N`, it is of course possible to do more complex experiments by wrapping the function in a loop and passing different values of other parameters, such as `treat_assign`.

